---
title: 'Introduction to the npsp Package'
author: 'Ruben Fernandez-Casal (ruben.fcasal@udc.es)'
date: '`r paste("npsp", packageVersion("npsp"))`'
output: 
  html_document:
    toc: yes
    toc_depth: 2
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to the npsp Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, fig.height=4, fig.width=7, 
                      fig.align = 'center', out.width = '100%')
# knitr::spin("npsp_intro2.R", knit = FALSE)
```

This vignette (a working draft) tries to illustrate the use of the `npsp` (Nonparametric spatial statistics) package. 
This package implements nonparametric methods
for inference on multidimensional geostatistical processes, avoiding the
misspecification problems that may arise when using parametric models.
The spatial process can be either stationary or show a non-constant
trend. Joint estimation of the trend and the semivariogram can be
performed automatically, by using the function `np.fitgeo()`, or by
a step-by-step approach.
    

# Introduction

We will assume that
$\left\{ Y(\mathbf{x}),\mathbf{x}\in D \subset \mathbb{R}^{d}\right\}$
is a spatial process that can be modeled as:
$$Y(\mathbf{x})=\mu(\mathbf{x})+\varepsilon(\mathbf{x}),$$ 
    
where $\mu(\cdot)$ is the trend function (large-scale variation)
and the error term $\varepsilon$ (small-scale variation), is a second order stationary
process with zero mean and covariogram
$C(\mathbf{u}) = Cov(\varepsilon \left( \mathbf{x}\right) ,\varepsilon \left(\mathbf{x}+\mathbf{u}\right) )$,
with $\mathbf{u} \in D$.

In this framework, given $n$ observed values
$\mathbf{Y}=( Y(\mathbf{x}_{1}),...,Y(\mathbf{x}_{n}))^{t}$, the
first step consists in estimate the trend $\mu(\mathbf{x})$ and the
semivariogram $\gamma(\mathbf{h}) = C(\mathbf{0}) - C(\mathbf{u})$.

There are several `R` packages, such as [`gstat`](https://r-spatial.github.io/gstat) or [`geoR`](http://www.leg.ufpr.br/geoR), which
implement the traditional geostatistical techniques to approximate
these functions. Nevertheless, as these approaches usually assume
parametric models, they can present misspecification problems.

The aim of the `npsp` package is to provide nonparametric tools for
geostatistical modeling and interpolation, under the general spatial
model (large + small variation scales), and without assuming any specific form
(parametric model) for the trend and the variogram of the process.


```{r }
library(npsp)
```


The `precipitation` data set (`sp::SpatialGridDataFrame` class object), supplied with the `npsp` package, will
be used in the examples in this work. The data consist of total
precipitations (square-root of rainfall inches) during March 2016
recorded over 1053 locations on the continental part of USA.


```{r }
library(sp)
summary(precipitation)
```



For instance, `spoints()` function may be used
to plot the data...


```{r }
spoints(precipitation)
```


Function `scattersplot()` may be used to detect the presence of a non-constant trend...

```{r scattersplot, fig.height=9, fig.width=9}
scattersplot(precipitation)
```


# Trend estimation

The local polynomial trend estimator
$\hat{\mu}_{\mathbf{H}}(\mathbf{x})$ (e.g. Fan and Gijbels, 1996), obtained by
polynomial smoothing of
$\{(\mathbf{x}_{i},Y(\mathbf{x}_{i})):i=1,\ldots,n\}$, is the
solution for $\beta_0$ to the least squares minimization problem
$$\begin{aligned}
    \min_{\beta_0 ,\boldsymbol{\beta}_1, \ldots, \boldsymbol{\beta}_p}
    \sum_{i=1}^{n}\left\{ Y(\mathbf{x}_{i})-\beta_0 
    -\boldsymbol{\beta}_1^{t}(\mathbf{x}_{i}-\mathbf{x}) - \ldots \right. \nonumber \\
    \left. -\boldsymbol{\beta}_p^{t}(\mathbf{x}_{i}-\mathbf{x})^p \right\}^{2}
    K_{\mathbf{H}}(\mathbf{x}_{i}-\mathbf{x}),
\end{aligned}$$ 
    where $\mathbf{H}$ is a $d\times d$ symmetric
positive definite matrix; $K$ is a $d$-dimensional kernel and
$K_{\mathbf{H}}(\mathbf{u})=\left\vert 
    \mathbf{H}\right\vert ^{-1}K(\mathbf{H}^{-1}\mathbf{u})$.

The bandwidth matrix $\mathbf{H}$ controls the shape and size of the
local neighborhood used to estimate $\mu(\mathbf{x})$.

The local trend estimates can be computed in practice with the
`locpol()` function. A full bandwidth matrix and a multiplicative
triweight kernel is used to compute the weights. Main calculations
are performed in Fortran using the LAPACK library.


```{r }
x <- coordinates(precipitation)
y <- precipitation$y
lp <- locpol(x, y, nbin = c(120, 120), h = diag(c(5, 5)))
```


We can plot the trend estimates with function `simage()` (or `spersp()`). 
Additionally, in this case the estimates outside of continental part of USA will be masked.

```{r }
attr <- attributes(precipitation)
border <- attr$border 
# Mask grid points outside of continental part of USA
lp <- mask(lp, window = border)
interior <- attr$interior 
slim <- range(y)
col <- jet.colors(256)
simage(lp,  slim = slim, main = "Trend estimates",
       col = col, xlab = "Longitude", ylab = "Latitude", asp = 1)
plot(border, border = "darkgray", lwd = 2, add = TRUE)
plot(interior, border = "lightgray", lwd = 1, add = TRUE)
```




The smoothing procedures in `npsp` use linear binning to discretize
the data. For instance, instead of the previous code we can use:

```{r }
bin <- binning(x, y, nbin = c(120, 120), window = border)
# lp <- locpol(bin, h = diag(c(5, 5)))
lp2 <- locpol(bin, h = diag(c(2, 2)))
simage(lp2, slim = slim, main = "Trend estimates",
       col = col, xlab = "Longitude", ylab = "Latitude", asp = 1)
plot(border, border = "darkgray", lwd = 2, add = TRUE)
plot(interior, border = "lightgray", lwd = 1, add = TRUE)
```


Usually two grids will be considered, a low resolution one
to speed up computations during the modeling (e.g. by using the
default values), and a higher resolution one to obtaing the final
results.

# Bandwidth selection

The trend estimates depends crucially on the bandwidth matrix $H$. 
A small bandwith will produce undersmoothed estimates, whereas a big bandwith will oversmooth the data. Although it may be subjectively choosed by eye, it could be very beneficial to have automatic selection criteria from the data.

Traditional bandwidth selectors, such as cross validation (CV) or
generalized cross validation (GCV), do not have a good performance
for dependent data (Opsomer et al, 2001), since they tend to undersmooth the trend
function.

The modified cross-validation criteria (MCV), proposed in (Chu and Marron, 1991) for time series, can be adapted for spatial data, by ignoring
observations in a neighbourhood $N(i)$ around $\mathbf{x}_{i}$. Note
that the ordinary CV approach is a particular case with
$N(i)=\left\lbrace \mathbf{x}_{i} \right\rbrace$.


In practice, the bandwidth can be selected through the `h.cv()` (or `hcv.data()`)
function.


```{r }
bin <- binning(x, y, window = border)
lp0.h <- h.cv(bin)$h
lp0 <- locpol(bin, h = lp0.h, hat.bin = TRUE) 
```


An alternative is the corrected generalized cross-validation
criterion (CGCV), proposed in (Francisco-Fernández and Opsomer, 2005), that takes the spatial
dependence into account. 
To use this approach in practice (by setting `objective = "GCV"` and `cov.bin` in `h.cv`), 
an estimate of the variogram is previously required.

# Variogram estimation

When the process is assumed stationary, the pilot local linear
estimate $\hat\gamma(\mathbf{u})$ is obtained by the linear
smoothing of
$\left(\mathbf{x}_{i}-\mathbf{x}_{j},(Y(\mathbf{x}_i)- Y(\mathbf{x}_j))^2/2 \right)$.
The corresponding bandwidth can be selected, for instance, by
minimizing the cross-validation relative squared error.

In the general setting with a non constant trend, the natural approach consists in removing
the trend and estimating the variogram from the residuals
$\mathbf{r}=\mathbf{Y}-\mathbf{SY}$. Nevertheless, the residuals
variability may be very different to that of the true errors (see
e.g. Cressie, 1993, Section 3.4.3, for the case of the linear trend
estimator). As the bias due to the direct use of residuals in
variogram estimation may have a significant impact on inference, a
similar approach to that described in Fernández-Casal and Francisco-Fernández (2014) could be considered.

In practice, the local linear variogram estimates can be computed
with functions `np.svar()` and `np.svariso.corr()`. The generic
function `h.cv()` may be used to select the corresponding bandwidth.


```{r }
svar.bin <- svariso(x, residuals(lp0), nlags = 60, maxlag = 20)  
svar.h <- h.cv(svar.bin)$h

svar.np <- np.svar(svar.bin, h = svar.h)
svar.np2 <- np.svariso.corr(lp0, nlags = 60, maxlag = 20, 
                            h = svar.h, plot = FALSE) 
```


The final variogram estimate is obtained by fitting a
"nonparametric" isotropic Shapiro-Botha variogram model (Shapiro and Botha, 1991), to
the nonparametric pilot estimate, by using the function
`fitsvar.sb.iso()`.


```{r }
svm0 <- fitsvar.sb.iso(svar.np, dk = 0) 
svm1 <- fitsvar.sb.iso(svar.np2, dk = 0) 
# plot...
plot(svm1, main = "Nonparametric bias-corrected semivariogram\nand fitted models", 
     legend = FALSE, xlim = c(0,max(coords(svar.np2))), 
     ylim = c(0,max(svar.np2$biny, na.rm = TRUE)))
plot(svm0, lwd = c(1, 1), add = TRUE)
plot(svar.np, type = "p", pch = 2, add = TRUE)
# abline(h = c(svm1$nugget, svm1$sill), lty = 3)
# abline(v = 0, lty = 3)
legend("bottomright", legend = c("corrected", 'biased'),
            lty = c(1, 1), pch = c(1, 2), lwd = c(1, 1))
```


# Automatic modeling

Function `np.fitgeo()` implements an automatic procedure for the
joint estimation of the trend and the semivariogram. The algorithm
is as follows:

1.  Select an initial bandwidth to estimate the trend using the MCV
    criterion (function `h.cv()`).

2.  Compute the local linear trend estimate (function `locpol()`).

3.  Obtain a bias-corrected pilot semivariogram estimate from the
    corresponding residuals (function `np.svariso.corr()`).

4.  Fit a valid Shapiro-Botha model to the pilot
    semivariogram estimates (function `fitsvar.sb.iso()`).

5.  Select a new bandwidth matrix for trend estimation with the CGCV
    criterion, 
    using the fitted variogram model to approximate the
    data covariance matrix
    (function `h.cv()` with `objective = "GCV"` and 
    `cov.bin =` fitted variogram model).

6.  Repeat steps 2-4 until convergence (in practice, only two
    iterations are usually needed).



```{r geomod, fig.height=4, fig.width=12}
geomod <- np.fitgeo(x, y, nbin = c(30, 30), maxlag = 20, svm.resid = TRUE, window = border)
plot(geomod)
```


Setting `corr.svar = TRUE` may be very slow (and memory demanding) when the number of data is large (note also that the bias in the residual variogram decreases when the sample size increases).


# Kriging

The final trend and variogram estimates could be employed for
spatial prediction, by using method `np.kriging()`.


```{r kriging, fig.height=4, fig.width=12}

krig.grid <- np.kriging(geomod, ngrid = c(120, 120))
# Plot kriging predictions and kriging standard deviations
old.par <- par(mfrow = c(1,2), omd = c(0.0, 0.98, 0.0, 1))
krig.grid2 <- krig.grid
simage(krig.grid2, 'kpred', main = 'Kriging predictions', slim = slim,
      xlab = "Longitude", ylab = "Latitude" , col = col, asp = 1)
plot(border, border = "darkgray", lwd = 2, add = TRUE)
plot(interior, border = "lightgray", lwd = 1, add = TRUE)
simage(krig.grid2, 'ksd', main = 'Kriging sd',
      xlab = "Longitude", ylab = "Latitude" , col = hot.colors(256), asp = 1)
plot(border, border = "darkgray", lwd = 2, add = TRUE)
plot(interior, border = "lightgray", lwd = 1, add = TRUE)
par(old.par)
```


Currently, only global residual kriging is implemented. Users are
encouraged to use `gstat::krige()` (or `krige.cv()`) together with
`as.vgm()` for local kriging 
(see e.g. https://rubenfcasal.github.io/npsp/articles/krigstat.html).


# Conclusions and future work

Unlike traditional geostatistical methods, the nonparametric
procedures avoid problems due to model misspecification.

The bias due to the direct use of residuals, may have a significant
impact on variogram estimation (and consequently on the selected
bandwidth with the CGCV criterion). Therefore, the use of the
bias-corrected nonparametric variogram estimator would be
recommended.

Currently, only isotropic semivariogram estimation is supported, but
the intention is to extend this approach to anisotropy in two
components (Fernández-Casal et al., 2003), 
which could be suitable for modeling spatio-temporal dependency.

The nonparametric trend and variogram estimates also allow to
perform inferences about other characteristics of interest of the
process, for instance, by using the bootstrap algorithm described in
(Castillo-Páez et al., 2019),
which will be implemented in the next version of the package
(https://github.com/rubenfcasal/npsp).


# Acknowledgments

This research has been supported by MINECO grant MTM2017-82724-R, and by the Xunta de Galicia (Grupos de Referencia Competitiva ED431C-2020-14 and Centro de Investigación del Sistema universitario de Galicia ED431G 2019/01), all of them through the ERDF.

# References

Castillo-Páez, S., Fernández-Casal, R., García-Soidán, P. (2019). A
nonparametric bootstrap method for spatial data. *Comput. Stat. Data
An.*, **137**, 1--15.

Chu, C. K. and Marron, J. S. (1991). Comparison of Two Bandwidth
Selectors with Dependent Errors. *The Annals of Statistics* **19**,
1906--1918.

Cressie, N. (1993). *Statistics for Spatial Data*. Wiley, New York.

Fernández-Casal, R. (2019). npsp: Nonparametric Spatial Statistics. R
package version 0.7-5. *http://github.com/rubenfcasal/npsp*.

Fan, J. and Gijbels, I. (1996). *Local polynomial modelling and its
applications*. Chapman & Hall, London.

Fernández-Casal, R. and Francisco-Fernández, M. (2014). Nonparametric
bias-corrected variogram estimation under non-constant trend.
*Stochastic Environmental Research and Risk Assessment* **28**,
1247--1259.

Fernández-Casal, R., González Manteiga, W. and Febrero-Bande, M. (2003).
Flexible Spatio-Temporal Stationary Variogram Models. *Statistics and
Computing* **13**, 127--136.

Francisco-Fernández, M. and Opsomer, J. D. (2005). Smoothing parameter
selection methods for nonparametric regression with spatially correlated
errors. *The Canadian Journal of Statistics* **33**, 279--295.

Opsomer, J. D., Wang, Y. and Yang, Y. (2001). Nonparametric regression
with correlated errors. *Statistical Science* **16**, 134--153.

Shapiro, A. and Botha, J.D. (1991). Variogram fitting with a general
class of conditionally non-negative definite functions. *Computational
Statistics and Data Analysis* **11**, 87--96.
